---
title: "Converting vcfR to other types"
author: "Brian J. Knaus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


Once we have finished examining our data in vcfR we'll want to format it so that other softwares can utilize it.
An straightforward path is to create a vcf.gz format file.
One downside to this path is that it creates an intermediate file.
When working on large datasets this intermediate file may be rather large.
If your path remains in R it may be preferable to convert your vcfR objects to objects defined by other packages.
Here we explore examples of these paths


## Data import

As in other vignettes, we begin by loading the example data.

```{r}
library(vcfR)
vcf_file <- system.file("extdata", "pinf_sc1_100_sub.vcf.gz", package = "vcfR")
vcf <- read.vcf(vcf_file, verbose = FALSE)
```



## Creating *.vcf.gz format files.


The function `write.vcf` can be used to create vcf.gz format files (gzipped vcf files) from vcfR objects (either objects of class vcfR or Chrom).


```{r}
original_dir <- getwd()
test_dir <- tempdir()
write.vcf(vcf, "test.vcf.gz")
unlink("test.vcf.gz") # Clean up after our example is done.
setwd(original_dir)
```


## Creating genind objects

The packages **adegenet** and **poppr** use objects of class genind.
We can create genind objects fairly easily.
First we'll need to extract a matrix of genotypes from the vcfR object (this works on Chrom objects as well).
Then we'll use an adegenet function to create the genind object.


```{r, eval=TRUE}
gt <- extract.gt(vcf)
my_genind <- adegenet::df2genind(t(gt), sep="/")
my_genind

class(my_genind)
```


Note that we need to transpose the matrix of genotypes (using `t()`).
This is because the VCF format stores samples in columns and loci in rows.
The `genind` definition expects samples in rows and loci in columns.


## Creating genclone objects


The package **poppr** uses objects of class genclone as well as genind.
Once a genind object has been created it is fairly straight forward to create a genclone object.


```{r, eval=FALSE}
my_genclone <- poppr::as.genclone(my_genind)
my_genclone
class(my_genclone)
```




## Creating genlight objects

To do.


## Creating loci objects

The package **pegas** uses objects of class loci.
Here we create a short script to help convert our matrix of genotypes to an object of class loci.


```{r}
vcfR2loci <- function(x, ...)
{
    if(class(x) == "Chrom")
    {
      x <- x@vcf
    }
    x <- extract.gt(x)
    # modified from pegas::as.loci.genind
    obj <- as.data.frame(t(x))
    icol <- 1:ncol(obj)
    for (i in icol) obj[, i] <- factor(obj[, i] )
    class(obj) <- c("loci", "data.frame")
    attr(obj, "locicol") <- icol
    obj
}

my_loci <- vcfR2loci(vcf)
class(my_loci)
```






