---
title: "Converting vcfR to other types"
author: "Brian J. Knaus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


Once we have finished examining our data in vcfR we'll want to format it so that other softwares can utilize it.
An straightforward path is to create a vcf.gz format file.
One downside to this path is that it creates an intermediate file.
When working on large datasets this intermediate file may be rather large.
If your path remains in R it may be preferable to convert your vcfR objects to objects defined by other packages.
Here we explore examples of these paths


## Data import

As in other vignettes, we begin by loading the example data.

```{r}
library(vcfR)
vcf_file <- system.file("extdata", "pinf_sc1_100_sub.vcf.gz", package = "vcfR")
vcf <- read.vcf(vcf_file, verbose = FALSE)
```



## Creating *.vcf.gz format files.


The function `write.vcf` can be used to create vcf.gz format files (gzipped vcf files) from vcfR objects (either objects of class vcfR or Chrom).


```{r}
original_dir <- getwd()
test_dir <- tempdir()
write.vcf(vcf, "test.vcf.gz")
unlink("test.vcf.gz") # Clean up after our example is done.
setwd(original_dir)
```


## Creating genind objects

The packages **adegenet** and **poppr** use objects of class genind.
We can create genind objects with the function vcfR2genind.


```{r, eval=TRUE}
my_genind <- vcfR2genind(vcf)
my_genind
adegenet::genind2df(my_genind, sep="/")[1:6, 1:8]
class(my_genind)
```


The function vcfR2genind calls extract.gt to create a matrix of genotypes.
This matrix is converted into a genind object with the adegenet function df2genind.


## Creating genclone objects


The package **poppr** uses objects of class genclone as well as genind.
Once a genind object has been created it is fairly straight forward to create a genclone object.


```{r, eval=FALSE}
my_genclone <- poppr::as.genclone(my_genind)
my_genclone
class(my_genclone)
```




## Creating genlight objects

To do.


## Creating loci objects

The package **pegas** uses objects of class loci.
We can use the function vcfR2loci to convert our vcfR object to one of class loci.


```{r}
my_loci <- vcfR2loci(vcf)
class(my_loci)
```






