---
title: "Filtering data with vcfR"
author: "Brian J. Knaus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering data with vcfR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


Most variant calling pipelines result in files containing variant information.
These files are typically intended to be filtered as an attempt to remove false positives, or otherwise problematic sites.
The R package vcfR provides tools to facilitate this filtering as well as to visualize the effects of choices made during this process.


## Data input


The input of data was covered in the vignette 'Introduction to vcfR.
If you haven't read this vignette, you can with:

```{r, eval=FALSE}
vignette("intro_to_vcfR", package="vcfR")
```

.

For this vignette, we'll use the example data provided with vcfR.


```{r}
library(vcfR)
data(vcfR_example)
pinf_mt <- create_chrom(name='pinf_mt', vcf=pinf_vcf, seq=pinf_dna, ann=pinf_gff)
pinf_mt <- proc_chrom(pinf_mt, verbose=FALSE)
```


We can now visualize this data.


```{r, fig.height=7, fig.width=7}
chromoqc(pinf_mt)
```



## Getting to know your vcf data




