---
title: "Visualizing vcf data I"
author: "Brian Knaus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization I}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


Most variant calling pipelines result in files containing variant information.
The [vcf format](http://samtools.github.io/hts-specs/ "vcf format at hts-specs") is an increasingly popular format for this data.
The format of these files and their content is discussed in the vignette 'vcf data'.
These files are typically intended to be post-processed (i.e., filtered) as an attempt to remove false positives or otherwise problematic sites.
The R package vcfR provides tools to facilitate this filtering as well as to visualize the effects of choices made during this process.


## Data input


The input of data was covered in the vignette 'Introduction to vcfR.
If you haven't read this vignette, you can with:

```{r, eval=FALSE}
vignette("intro_to_vcfR", package="vcfR")
```



For this vignette, we'll use the example data provided with vcfR.


```{r}
library(vcfR)
vcf_file <- system.file("extdata", "pinf_sc1_100_sub.vcf.gz", package = "vcfR")
seq_file <- system.file("extdata", "pinf_sc100.fasta", package = "vcfR")
gff_file <- system.file("extdata", "pinf_sc100.gff", package = "vcfR")

vcf <- read.vcf(vcf_file, verbose = FALSE)
dna <- ape::read.dna(seq_file, format = "fasta")
gff <- read.table(gff_file, sep="\t")

chrom <- create.chromR(name="Supercontig_1.100", vcf=vcf, seq=dna, ann=gff, verbose=TRUE)
chrom <- masker(chrom, min_DP = 300, max_DP = 700)
chrom <- proc.chromR(chrom, verbose = TRUE)

```


The function create.chromR() reports the names of the chromosome (supercontig, contig, etc.) so that we can verify that all three data sources contain the same chromosome.


We can now visualize this data.


```{r, fig.height=7, fig.width=7}
chromoqc(chrom)
```


In vcfR we refer to this form of plot as a **chromo plot**.
This plot summarizes data from our three data sources.
Sequence data is represented on the lowest plot.
Here called nucleotides are represented in green while uncalled nucleotides (Ns) appear in red.
This chromosome contains all called nucleotides.
Depending on the quality of your genome, it may include streches of uncalled nucleotides.
Above the sequence is an annotation track.
Features in the annotation file (e.g., gene models, exons, etc.) are represented as dark red rectangles.
Above the annotation track is a windowing analysis of G/C content.
This track includes marginal barplots to summarize its contents, as do the tracks above it.
The next track summarizes the number of variants per window.
Above that we see quality, mapping quality and read depth.


One of the strengths of R is that it is interactively implemented.
This means we can change things fairly rapidly.
For example, the above example used 3 kbp windows.
If we become concerned with how our choice of window size affects our analysis, we can change it.



```{r, fig.height=7, fig.width=7}
chrom <- proc.chromR(chrom, verbose=FALSE, win.size=1e3)
chromoqc(chrom, verbose=FALSE)
```





We can also use the function head() to summarize the contents of the Chrom object.


```{r}
head(chrom)
```


The slots of the class Chrom are defined in the package documentation.


## Getting to know your vcf data


The vcf format is discussed in the vignette 'vcf data'.
In the Chrom object, this information occurs as the slots 'vcf.meta', 'vcf.fix' and 'vcf.gt'.
Some of this information can be rapidly explored using the plot function.


```{r, fig.height=7, fig.width=7}
plot(chrom)
```


Here we see the distribution of some quality metrics.
The raw read depth and mapping quality have been extracted from the INFO column of the vcf data.
The quality is from the QUAL column of the vcf data.
And the variant count per window was summarized during the windowing analysis performed by proc.chromR().
The raw read depth appears fairly continuous, an observation which appears consistent with the above chromo plots.
Mapping quality appears to be predominantly one value (about 45), this may not have been apparent in teh chromo plots.
The quality jumps out as bimodal, suggesting it may be a good candidate for filtering.
And the varaint count per window summarizes the degree of variability we've observed along our chromosome, and suggest we may have some highly variable regions.



Comparing the chomo plots with the histograms we may wonder if mapping quality is really being driven by raw read depth.
Thsi idea can be explored with a scatterplot.



```{r, fig.height=7, fig.width=7}
plot(chrom@var.info$DP, chrom@var.info$MQ, col="#0000ff66")
```


From this perspective, there does not appear to be a strong correlation.
This can be validated by calculating a correlation.



```{r}
cor(chrom@var.info$DP, chrom@var.info$MQ)
```


